Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students

### 2024-02-28 demystify.tex
updates

Most universities the US have CS programs. 
However, while many universities offer PhD in their CS programs, many do not.  These universities might offer just Bachelor degrees (e.g., BS) and no graduate studies (i.e., no MS or Ph.D degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has Ph.D. in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiple locations of the University of Texas, e.g., Austin, Dallas, Arlington, have PhDs in CS.

@@

  \item \emph{No faculty available to advise you}.  You could have an excellent profile, but if there is no faculty who can advise you (e.g., they already have too many students), you will not be admitted. This is actually for your own good because you do not want to be in a program where you cannot find an adviser.
\end{itemize}

\section{Should I ask for feedback?}
No, don't bother.  You will likely not get any useful feedback.  We cannot reveal your evaluation results (e.g., if we say because you have weak LORs, then that would reveal something about your writers) or give you advice on how to improve. So it is best to just move on.  If you really want feedback, ask your professors, collaborators, ref writers, or those who have previously applied.

@@

One main concern for non-CS or non-STEM students is if you have the sufficient technical background, typically obtained through core CS courses.  So you will want to show that you have such knowledge through your coursework, projects, or research. 
For example, if you have taken a course on Algorithms, even online ones like Coursera, you can talk about it in your statement.  If you have done a project that requires knowledge of OS or have a professional certification (e.g., A+) through work, you can talk about it in your statement.  If you have done research that requires knowledge of Discrete Maths, you can talk about it in your statement.  You can also ask your LoR writers to talk about your technical background.  
In summary, in your application, convince us that you have the technical background to do CS PhD research. 
Note that many online courses teaching about AI does not really help much as these are not core CS knowledge.

In short, you \emph{do not need} to formally taking CS courses, you just need to show that you have these essential knowledge through ways as mentioned. Many universities are well-aware that incoming graduate students might not have all the technical background, so they often have a "bridge" courses to help students catch up.  For example, GMU has four bridge courses (Data Structures and Algorithms, Computer Systems, Discrete Math, and Programming Foundation) that incoming students can take to catch up on their CS knowledge.

@@

\section{Is an MS degree required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, student with BS can get MS degree ``along the way'' to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.

>>>>

Most universities the US have a CS program.
However, while many universities offer PhD in their CS programs, many do not.  These universities might offer just Bachelor degrees (e.g., BS) and no graduate studies (i.e., no MS or Ph.D degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has Ph.D. in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiple locations of the University of Texas, e.g., Austin, Dallas, Arlington, have PhDs in CS.

@@

  \item \emph{No faculty available to advise you}.  You could have an excellent profile, but if there is no faculty who is willing to advise you (e.g., they already have too many students), you will not be admitted. This is actually for your own good because you do not want to be in a program where you cannot find an adviser.
\end{itemize}

\paragraph{Should I ask for feedback?}
No, don't bother.  You will likely not get any useful feedback.  We cannot reveal your evaluation results (e.g., if we say because you have weak LORs, then that would reveal something about your writers) or give you advice on how to improve. So it is best to just move on.  If you really want feedback, ask your professors, collaborators, ref writers, or those who have previously applied.

@@

One main concern for non-CS or non-STEM students is if you have the sufficient technical background, typically obtained through core CS courses.  You will want to show that you have such knowledge through your coursework, projects, or research.
For example, if you have taken a course on Algorithms, even online ones like Coursera, you can talk about it in your statement.  If you have done a project that requires knowledge of OS or have a professional certification (e.g., A+) through work, you can talk about it in your statement.  If you have done research that requires knowledge of Discrete Maths, you can talk about it in your statement.  You can also ask your LoR writers to talk about your technical background.  
In summary, in your application, convince us that you have the technical background to do CS PhD research. 
Note that many online courses teaching about AI does not really help much as these are not core CS knowledge.

In short, you \emph{do not need} to formally taking CS courses, you just need to show that you have these essential knowledge, e.g., through the mentioned ways. Many universities are well-aware that incoming graduate students might not have all the technical background, so they often have a "bridge" courses to help students catch up.  For example, GMU has four bridge courses (Data Structures and Algorithms, Computer Systems, Discrete Math, and Programming Foundation) that incoming students can take to catch up on their CS knowledge.

@@

\section{Is an MS required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, student with BS can get MS degree ``along the way'' to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.
